---
title: "STAT574E Group 3 Final Project Appendix"
author: "Alex Salce, Raymond Owino, Matthew Wallace"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    theme: cosmo
  pdf_document: default
header-includes:
- \renewcommand*\familydefault{\sfdefault} %% this picks a sans serif font
- \usepackage[T1]{fontenc}
---

```{r warning=FALSE, message=FALSE, echo=FALSE}

library(dplyr)
library(tigris)
library(sf)
library(ggplot2)
library(mapview)
library(tidyverse)
library(scales)
library(knitr)
library(readr)

```

Packages we need.

```{r datasetload, warning=FALSE, message=FALSE, echo=FALSE}

# load datasets
library(here)
set_here(path='../../')

load(here('data/wfigs_az_sf_EPSG32612.RData'))
load(here('data/az_rd_4WD_EPSG32612.RData'))
load(here('data/az_rd_primary_EPSG32612.RData'))
load(here('data/az_rd_secondary_EPSG32612.RData'))


#NOTE: all sf object data uses CRS ESPG:32612
# https://epsg.io/32612


```

We have a few datasets that we have pre-constructed based on available resources and challenges with acquiring data for the features that we are interested in modeling. Note that all `sf` objects are projected to `EPSG:3612` ([https://epsg.io/32612](https://epsg.io/32612)).

- `wfigs_az_sf` is a `sf` object with 18089 observations of wildfire incidence in the state of Arizona. This data was originally acquired via [Wildland Fire Incident Locations](https://data-nifc.opendata.arcgis.com/datasets/nifc::wildland-fire-incident-locations/about) from the National Interagency Fire Center. It contains spatial point data indicating the origin of each wildfire recorded in the [IRWIN](https://www.wildfire.gov/application/irwin-integrated-reporting-wildfire-information) database, and includes many useful features. Most notably, the data includes the `IncidentSize`, which is the size in acres of the resulting wildfire. `IncidentSize` is the primary response variable of interest for our study. The data included in this dataframe includes hand-selected features that were deemed to be potentially useful during the exploration phase of the project, as well as other data manually recovered for natural and environmental factors, distances to roads in Arizona, as well as census data for population density, discussed elsewhere in the report.

-`az_rd_primary` and `az_rd_secondary` are `sf` data of all roads in Arizona with FCC road classification codes of `S1100` or `S1200`. These two classes are treated as "major" roads in our analyses. `az_rd_4WD` is all roads with `S1500` classification, and are representative of remote roadways in Arizona. All other road types are removed/disregarded as they are deemed not useful for our analysis. [See this document for all classification codes](https://docs.fcc.gov/public/attachments/DA-12-187A3.pdf).


```{R azstatesf, cache = TRUE, message = FALSE, warning = FALSE, results = 'hide'}

arizona_sf <- states() %>% filter_state("arizona")
arizona_sf <- st_transform(arizona_sf, crs = "EPSG:32612")

az_counties_sf <- counties(state = "AZ", cb = TRUE)
az_counties_sf <- st_transform(az_counties_sf, crs = "EPSG:32612")

```
First, we are going to load the `sf` object for the state of Arizona and project to `EPSG:32612`.

### Create Wildfire Dataframes

```{r finaldataframes}

az_fires_rx <- wfigs_az_sf %>% filter(IncidentTypeCategory=="RX")
az_fires_wf <- wfigs_az_sf %>% filter(IncidentTypeCategory=="WF")

az_fires_wf_hum <- az_fires_wf %>% filter(FireCause=="Human")
az_fires_wf_nat <- az_fires_wf %>% filter(FireCause=="Natural")
az_fires_wf_un <- az_fires_wf %>% filter(FireCause=="Undetermined" | FireCause=="Unknown")


```

For our analyses using `wfigs_az_sf`, we want to be sure to differentiate between fires that are prescribed burns (`IncidentTypeCategory=="RX"`) and wildfires (IncidentTypeCategory=="WF"). Our analyses will only use fires of type `WF`, since prescribed burns are fires deliberately started and controlled by a service entity.

We also want to differentiate fires by their causes. There are four unique categories of fire in the `FireCause` column, `HUMAN`, `NATURAL`, `UNDETERMINED`, and `UNKNOWN`. For our analyses, we will discard `UNDETERMINED` and `UNKNOWN` type fires, as we cannot reasonably assume anything about them. Additionally, they comprise only about 12% of all of the `wfigs_az_sf` data.


## Histograms

### Nearest Roads

This histogram uses input `sf` object to create `histdf`, which is used to generate a histogram of the number of closest roads, either primary/secondary roads or 4wd roads, to each wildfire incident origin. The $x$-axis indicates the distance in meters of that bin to wildfire points.

As a generalization, we will think of Primary & Secondary roads as essentially the same class of roads, and we know that many humans use these roads every day. No matter the stretch of road, there is a good chance of there being some human settlement of some kind alongside these roads within close proximity.

The "4WD" roads are technically a fairly "incomplete" dataset, but they do give a sense of areas humans can spend their time in remote areas. In general, though, if a wildfire incident is closer to a 4WD road we can think of it as being a "more remote" fire, i.e. there is a good chance it's fairly removed from any prominent human settlement.

For reference, the minimum distances as well as the closest road type indicators were generated using the below code.

```{r nearestrdDataGen, eval=FALSE}

wfigs_az_sf$distance_rd_primary <-
  st_distance(wfigs_az_sf, az_rd_primary) %>% apply(1, min)

wfigs_az_sf$distance_rd_secondary <-
  st_distance(wfigs_az_sf, az_rd_secondary) %>% apply(1, min)

wfigs_az_sf$distance_rd_4wd <-
  st_distance(wfigs_az_sf, az_rd_4wd) %>% apply(1, min)


wfigs_az_sf <- wfigs_az_sf %>%
  mutate(distance_rd_min_prisec = pmin(distance_rd_primary, 
                                    distance_rd_secondary))

wfigs_az_sf <- wfigs_az_sf %>%
  mutate(distance_rd_min_all = pmin(distance_rd_primary, 
                                    distance_rd_secondary,
                                    distance_rd_4wd))

wfigs_az_sf$distance_rd_min_isprisec <- as.integer(wfigs_az_sf$distance_rd_min_all ==
                                                     wfigs_az_sf$distance_rd_min_prisec) 

```

#### All wildfire data

We will first look at the "Nearest Roads" histograms for all wildfire data (no filtering for the size of the fire.)

```{r disthistsmindistprimsec4wdNothresh, echo=FALSE}


bins <- 30

# human caused

histdf1 <- az_fires_wf_hum

# naturally caused

histdf2 <- az_fires_wf_nat

# undetermined / unknown cause

histdf3 <- az_fires_wf_un


x_rng <- range(c(histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==1], 
                 histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==0],
                 histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==1], 
                 histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==0]))
y_max <- max(hist(histdf1$distance_rd_min_all, plot = FALSE)$counts, 
             hist(histdf2$distance_rd_min_all, plot = FALSE)$counts)


#overlay hist of dist from roads

# one plot
# par(mfrow=c(1, 2))

hist(histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==1], 
     col = rgb(0, 0, 1, 0.5), xlim = x_rng, ylim = c(0, y_max),
     breaks = bins,
     main = "Min distance to Human caused fires", xlab = "Distance (m)")
hist(histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==0], 
     col = rgb(1, 0, 0, 0.5), 
     breaks = bins,
     add = TRUE)

# Add a legend
legend("topright", c("MinDistIsPrimSec", "MinDistIs4WD"),
       fill = c(rgb(0, 0, 1, 0.5), rgb(1, 0, 0, 0.5)))

hist(histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==1], 
     col = rgb(0, 0, 1, 0.5), xlim = x_rng, ylim = c(0, y_max),
     breaks = bins,
     main = "Min distance to naturally caused fires", xlab = "Distance (m)")
hist(histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==0], 
     col = rgb(1, 0, 0, 0.5), 
     breaks = bins,
     add = TRUE)

# Add a legend
legend("topright", c("MinDistIsPrimSec", "MinDistIs4WD"),
       fill = c(rgb(0, 0, 1, 0.5), rgb(1, 0, 0, 0.5)))



df_names <- c("Human Caused Fires", "Naturally Caused Fires", "Unknown Caused Fires")
row_counts <- c(nrow(az_fires_wf_hum), nrow(az_fires_wf_nat), nrow(az_fires_wf_un))

# Create a table
result_table <- data.frame(WildFireType = df_names, Counts = row_counts)

kable(result_table, label = "Wildfire Counts")

rm(histdf1, histdf2, histdf3, x_rng, y_max, bins, df_names, row_counts, result_table)

```

There is a very high concentration of wildfires cause by humans very close to primary and secondary roads, and natural fires generally seem to be more remote (in general closer to more 4WD roads.)


#### Thresholded for Wildfire Acreage

We define the threshold for a fire to be "large" as `IncidentSize` $\geq 1000$ acres. We threshold the data accordingly and replot our histograms.

```{r thresholding_1}

####################################
# RUN BELOW FOR FIRE SIZE THRESHOLD#
####################################


# fire size threshold
# FIRE_SIZE_CLASS = Code for fire size based on the number of acres within the 
# final fire perimeter (A=greater than 0 but less than or equal to 0.25 acres, 
# B=0.26-9.9 acres, C=10.0-99.9 acres, D=100-299 acres, E=300 to 999 acres, 
# F=1000 to 4999 acres, and G=5000+ acres).

# class F and G fires
wf_size_threshold <- 1000

# max prescribed burn size
# wf_size_threshold <- max(az_fires_rx$IncidentSize)

wfigs_az_sf$size_threshold <- as.integer(wfigs_az_sf$IncidentSize >= wf_size_threshold)

wfigs_az_sf_thresh <- wfigs_az_sf %>% filter(size_threshold==1)
# wfigs_az_sf_thresh <- wfigs_az_sf %>% filter(IncidentSize >= wf_size_threshold)


az_fires_rx <- wfigs_az_sf_thresh %>% filter(IncidentTypeCategory=="RX")
az_fires_wf <- wfigs_az_sf_thresh %>% filter(IncidentTypeCategory=="WF")


#human caused wildfires

table(wfigs_az_sf_thresh$FireCause)

az_fires_wf_hum <- az_fires_wf %>% filter(FireCause=="Human")
az_fires_wf_nat <- az_fires_wf %>% filter(FireCause=="Natural")
az_fires_wf_un <- az_fires_wf %>% filter(FireCause=="Undetermined" | FireCause=="Unknown")

```

```{r ppphist, echo=FALSE}

bins <- 30

# human caused

histdf1 <- az_fires_wf_hum

# naturally caused

histdf2 <- az_fires_wf_nat

# undetermined / unknown cause

histdf3 <- az_fires_wf_un


x_rng <- range(c(histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==1], 
                 histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==0],
                 histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==1], 
                 histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==0]))
y_max <- max(hist(histdf1$distance_rd_min_all, plot = FALSE)$counts, 
             hist(histdf2$distance_rd_min_all, plot = FALSE)$counts)


#overlay hist of dist from roads

# one plot
# par(mfrow=c(1, 2))

hist(histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==1], 
     col = rgb(0, 0, 1, 0.5), xlim = x_rng, ylim = c(0, y_max),
     breaks = bins,
     main = "Min distance to Human caused fires", xlab = "Distance (m)")
hist(histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==0], 
     col = rgb(1, 0, 0, 0.5), 
     breaks = bins,
     add = TRUE)

# Add a legend
legend("topright", c("MinDistIsPrimSec", "MinDistIs4WD"),
       fill = c(rgb(0, 0, 1, 0.5), rgb(1, 0, 0, 0.5)))

hist(histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==1], 
     col = rgb(0, 0, 1, 0.5), xlim = x_rng, ylim = c(0, y_max),
     breaks = bins,
     main = "Min distance to naturally caused fires", xlab = "Distance (m)")
hist(histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==0], 
     col = rgb(1, 0, 0, 0.5), 
     breaks = bins,
     add = TRUE)

# Add a legend
legend("topright", c("MinDistIsPrimSec", "MinDistIs4WD"),
       fill = c(rgb(0, 0, 1, 0.5), rgb(1, 0, 0, 0.5)))

df_names <- c("Human Caused Fires", "Naturally Caused Fires", "Unknown Caused Fires")
row_counts <- c(nrow(az_fires_wf_hum), nrow(az_fires_wf_nat), nrow(az_fires_wf_un))

# Create a table
result_table <- data.frame(WildFireType = df_names, Counts = row_counts)

kable(result_table, label = "Wildfire Counts")

rm(histdf1, histdf2, histdf3, x_rng, y_max, bins, df_names, row_counts, result_table)

```



## Models

### Spatial Linear Model - Raymond

A spatial linear model, capturing factors as well as spatial variation.


### Point Process Spatial Intensity Model - Matthew

Checking K, F, G functions to determine if wildfires are CSR or not.


### Point Process Binary GLM - Alex

#### Threshold for large wildfires

```{r thresholding_2}

####################################
# RUN BELOW FOR FIRE SIZE THRESHOLD#
####################################


# fire size threshold
# FIRE_SIZE_CLASS = Code for fire size based on the number of acres within the 
# final fire perimeter (A=greater than 0 but less than or equal to 0.25 acres, 
# B=0.26-9.9 acres, C=10.0-99.9 acres, D=100-299 acres, E=300 to 999 acres, 
# F=1000 to 4999 acres, and G=5000+ acres).

# class F and G fires
wf_size_threshold <- 1000

# max prescribed burn size
# wf_size_threshold <- max(az_fires_rx$IncidentSize)

wfigs_az_sf$size_threshold <- as.integer(wfigs_az_sf$IncidentSize >= wf_size_threshold)

wfigs_az_sf_thresh <- wfigs_az_sf %>% filter(size_threshold==1)
# wfigs_az_sf_thresh <- wfigs_az_sf %>% filter(IncidentSize >= wf_size_threshold)


az_fires_rx <- wfigs_az_sf_thresh %>% filter(IncidentTypeCategory=="RX")
az_fires_wf <- wfigs_az_sf_thresh %>% filter(IncidentTypeCategory=="WF")


#human caused wildfires

table(wfigs_az_sf_thresh$FireCause)

az_fires_wf_hum <- az_fires_wf %>% filter(FireCause=="Human")
az_fires_wf_nat <- az_fires_wf %>% filter(FireCause=="Natural")
az_fires_wf_un <- az_fires_wf %>% filter(FireCause=="Undetermined" | FireCause=="Unknown")

```
The first step is to filter our data to only "large" wildfires (`IncidentSize` $\geq 1000$ acres). We are now treating our data as point process data, so the magnitude of the resulting wildfire is no longer of interest. Rather, we will be only studying the occurrence of "big" wildfires.

Since we will be treating the wildfires as a point process, the most sensible data to study should be the naturally occurring fires as it would be reasonable to assume that they occur randomly. For example, naturally occurring fires can be the result of lightning strikes, which we have previously seen modeled using a point process approach.


#### Counties

```{r firesbycounty}

# az_county_intersections <- st_intersection(az_counties_sf, rbind(az_fires_wf_hum, az_fires_wf_nat))

az_county_intersections <- st_intersection(az_counties_sf, az_fires_wf_nat)

county_counts <- az_county_intersections %>%
  group_by(NAME) %>%
  summarise(count = n())

az_counties_with_counts <- az_counties_sf %>%
  st_join(county_counts, by = "NAME") %>%
  mutate(count = replace_na(count, 0))


ggplot(az_counties_with_counts) +
  geom_sf(aes(fill = count)) +
  scale_fill_viridis_c(name = "Point Count") +
  theme_minimal() +
  labs(title = "Number of Points per County in Arizona")

rm(county_counts, az_counties_with_counts, az_county_intersections)

```

```{r cococounty}

# Extract Coconino County
coconino_sf <- az_counties_sf %>% 
  filter(NAME == "Coconino")

#filter large wildfire data to only Coconino County
az_fires_wf <- st_intersection(az_fires_wf[az_fires_wf$FireCause=="Human" | az_fires_wf$FireCause == "Natural",], coconino_sf)

```

For this portion of the study, we opted to start smaller by filtering the data down to only Coconino County. Coconino County has the most naturally occurring wildfires in the state by a good margin, and is home to part of the largest contiguous Ponderosa Pine forest in the United States. 

#### Model

##### Additional Packages

```{r sglmlibrary, message=FALSE, warning=FALSE}

library(spatstat)
library(spmodel)
library(lubridate)
library(tidycensus)
library(ggmap)

```

##### Data Refinement & Additional Exploration

```{r refreshfilter, echo=FALSE}

az_fires_wf_hum <- az_fires_wf %>% filter(FireCause=="Human")
az_fires_wf_nat <- az_fires_wf %>% filter(FireCause=="Natural")
az_fires_wf_un <- az_fires_wf %>% filter(FireCause=="Undetermined" | FireCause=="Unknown")

```

```{r disthistsmindistprimsec4wdThresh, echo=FALSE}

bins <- 10

# human caused

histdf1 <- az_fires_wf_hum

# naturally caused

histdf2 <- az_fires_wf_nat

# undetermined / unknown cause

histdf3 <- az_fires_wf_un


x_rng <- range(c(histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==1], 
                 histdf1$distance_rd_min_all[histdf1$distance_rd_min_isprisec==0],
                 histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==1], 
                 histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==0]))
y_max <- max(hist(histdf1$distance_rd_min_all, plot = FALSE)$counts, 
             hist(histdf2$distance_rd_min_all, plot = FALSE)$counts)


#overlay hist of dist from roads


hist(histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==1], 
     col = rgb(0, 0, 1, 0.5), xlim = x_rng, ylim = c(0, y_max),
     breaks = bins,
     main = "Min distance to naturally caused fires", xlab = "Distance (m)")
hist(histdf2$distance_rd_min_all[histdf2$distance_rd_min_isprisec==0], 
     col = rgb(1, 0, 0, 0.5), 
     breaks = bins,
     add = TRUE)

# Add a legend
legend("topright", c("MinDistIsPrimSec", "MinDistIs4WD"),
       fill = c(rgb(0, 0, 1, 0.5), rgb(1, 0, 0, 0.5)))

df_names <- c("Human Caused Fires", "Naturally Caused Fires", "Unknown Caused Fires")
row_counts <- c(nrow(az_fires_wf_hum), nrow(az_fires_wf_nat), nrow(az_fires_wf_un))

# Create a table
result_table <- data.frame(WildFireType = df_names, Counts = row_counts)

kable(result_table, label = "Wildfire Counts")

rm(histdf1, histdf2, histdf3, x_rng, y_max, bins, df_names, row_counts, result_table)

```


##### Building a Background Poisson Process Realization

```{r builddataspglmppp, cache=TRUE, warning=FALSE, message=FALSE}

# set seed to 219 for report!
set.seed(219)

# set seed to 20 for predictions!
# set.seed(20)

# pick df of interest

az_fires_wf_spglm <- az_fires_wf_nat


az_fires_wf_spglm_ppp <- as.ppp(az_fires_wf_spglm)

background <- rpoispp((az_fires_wf_spglm_ppp$n) / area(as.owin(az_fires_wf_spglm_ppp)),
                      win = as.owin(coconino_sf))

df <- data.frame(x = background$x, y = background$y)

background_sf <- st_as_sf(df, coords = c("x", "y"), crs = "EPSG:32612")

```

```{r azccplotppp}

wf_sf_cc_plot <- st_as_sf(as.data.frame(az_fires_wf_spglm_ppp), coords = c("x", "y"), crs="EPSG:32612")

wf_sf_cc_plot <- st_transform(wf_sf_cc_plot, crs = 4326)

bkg_sf_cc_plot <- st_as_sf(as.data.frame(background), coords = c("x", "y"), crs="EPSG:32612")

bkg_sf_cc_plot <- st_transform(bkg_sf_cc_plot, crs = 4326)

coconino_plot <- coconino_sf %>% st_transform(4326)
az_cc_bbox <- st_bbox(coconino_plot)

az_cc_bbox <- c(
  left = az_cc_bbox["xmin"][[1]]-1,
  bottom = az_cc_bbox["ymin"][[1]]-1,
  right = az_cc_bbox["xmax"][[1]]+1,
  top = az_cc_bbox["ymax"][[1]]+1
)

coconino_map <- get_stadiamap(bbox = az_cc_bbox, zoom = 8)

# plot
ggmap(coconino_map) + labs(title = "Coconino County Naturally Caused Large Wildfire Incidence", subtitle = "Background realization marked with X") + theme_minimal() +
  geom_sf(data = coconino_plot, fill = NA, color = "black", size = 5, inherit.aes = FALSE) +
  geom_sf(data=wf_sf_cc_plot, shape = 21, size = 0.9, 
          color = "orangered3", fill = "orangered3", inherit.aes = FALSE) +
  geom_sf(data=bkg_sf_cc_plot, shape = 4, size = 1.1, color = "red4", inherit.aes = FALSE)

```

We first generate a realization of a Poisson point process to represent where Naturally occurring wildfires could have happened in Cocinino County.

```{r buildbkgcovs, warning=FALSE, message=FALSE, eval=FALSE}

## Build covariate data for background points

background_sf$distance_rd_primary <-
  st_distance(background_sf, az_rd_primary) %>% apply(1, min)

background_sf$distance_rd_secondary <-
  st_distance(background_sf, az_rd_secondary) %>% apply(1, min)

background_sf$distance_rd_4wd <-
  st_distance(background_sf, az_rd_4wd) %>% apply(1, min)


# dist_road covariates

background_sf <- background_sf %>%
  mutate(distance_rd_min_all = pmin(distance_rd_primary, 
                                    distance_rd_secondary, 
                                    distance_rd_4wd))

background_sf <- background_sf %>%
  mutate(distance_rd_min_prisec = pmin(distance_rd_primary, 
                                    distance_rd_secondary))

background_sf$distance_rd_min_isprisec <- as.integer(
  background_sf$distance_rd_min_all ==
    background_sf$distance_rd_min_prisec) 

# generate random dates

start_date <- ymd_hms("2014-01-01 00:00:00")
end_date <- now()
n <- nrow(background_sf)

random_dates <- sample(seq(start_date, end_date, by = "min"), n, replace = TRUE)
random_dates <- as.POSIXlt(random_dates)

background_sf$FireDiscoveryDateTime <- random_dates


## Population density data

population_data <- get_decennial(
  geography = "tract",
  variables = c(population="P001001"),
  state = "AZ",
  year = 2010,
  geometry = TRUE
)
population_data$variable = population_data$value/st_area(population_data$geometry)
names(population_data) <- c("GEOID","NAME","pop.density","pop.","geometry")

population_data <- st_transform(population_data, crs = "EPSG:32612")

background_sf <- st_join(background_sf, population_data, left = TRUE)

## Import Covariates

background_nat <- read_csv(here('data/background_final.csv'))

background_sf <- cbind(background_sf, background_nat[,11:19])


```

```{r importbkgcovs}

load(here('data/background_sf.RData'))

```


```{r gencsvforRaymond, eval=FALSE, echo=FALSE}

coords <- st_coordinates(background_sf)
df <- st_drop_geometry(background_sf)
df$x <- coords[, 1]
df$y <- coords[, 2]
write.csv(df, file = 'predictions_sf.csv', row.names = FALSE)

rm(df, coords)


```

```{r combine bkgwfcovariates}

df1 <- st_drop_geometry(az_fires_wf_spglm)
df2 <- st_drop_geometry(background_sf)


common_cols <- intersect(names(df1), names(df2))

df1 <- df1[, c(common_cols)]
df2 <- df2[, c(common_cols)]

df1$FireDiscoveryDateTime <- as.Date(df1$FireDiscoveryDateTime)
df2$FireDiscoveryDateTime <- as.Date(df2$FireDiscoveryDateTime)


Covariates <- rbind(df1, df2)

rm(df1,df2,df)

```


```{r}

all_points <- superimpose(unmark(az_fires_wf_spglm_ppp), background)

Wildfires <- c(rep(1, az_fires_wf_spglm_ppp$n), rep(0, background$n))

data <- cbind(Wildfires, Covariates)

data_sf <- st_as_sf(cbind(data, as.data.frame(all_points)[, c('x', 'y')]),
coords = c('x', 'y'))

```

```{r functions}

get_season <- function(date) {
  month <- as.integer(format(date, "%m"))
  season <- case_when(
    month %in% c(12, 1, 2) ~ 1,
    month %in% c(3, 4, 5) ~ 2,
    month %in% c(6, 7, 8) ~ 3,
    month %in% c(9, 10, 11) ~ 4
  )
  return(season)
}


```

```{r modelAIC, eval=FALSE}


# Create a list of formula objects
formula_list <- list(
  as.formula("Wildfires ~ I(sqrt(distance_rd_min_prisec)) +
  I(sqrt(distance_rd_4wd)) + I(sqrt(distance_rd_min_isprisec)) +
  I(log(mean_slope)) + mean_forest + mean_grass + I(log(pop.density)) +
  Precipitation_Buffered + Temp_Min_Buffered + Temp_Max_Buffered +
  I(month(FireDiscoveryDateTime))"),
  as.formula("Wildfires ~  I(sqrt(distance_rd_min_isprisec)) + pop.density +
  Precipitation_Buffered + Temp_Min_Buffered + I(get_season(FireDiscoveryDateTime)) +
  mean_grass * mean_forest"),
  as.formula("Wildfires ~  I(sqrt(distance_rd_min_isprisec)) + pop.density +
  Precipitation_Buffered + Temp_Min_Buffered + I(get_season(FireDiscoveryDateTime)) +
  mean_grass * mean_forest")
  # as.formula("Wildfires ~  I(sqrt(distance_rd_min_isprisec)) +
  # Precipitation_Buffered + Temp_Min_Buffered")
)

# Define spatial covariance types
# gaussian best for all so far
# OUT: cauchy and matern
# OUT/dont converge: wave and circular
cov_types <- c("wave", 
               "gaussian", 
               "spherical",
               "circular")

# Initialize a list to store results
results <- list()

# Outer loop for formulas
for (i in seq_along(formula_list)) {
  # Inner loop for spatial covariance types
  for (j in seq_along(cov_types)) {
    # Fit the model using spglm
    model <- spglm(formula_list[[i]], data = data_sf, family = binomial, spcov_initial = spcov_initial(cov_types[j]))
    
    # Store AIC in results list
    results[[paste0("Model_", i, "_", cov_types[j])]] <- AIC(model)
  }
}


```


```{r}

# best AIC model

spglm_formula <- Wildfires ~  I(sqrt(distance_rd_min_isprisec)) + pop.density +
  Precipitation_Buffered + Temp_Min_Buffered + I(get_season(FireDiscoveryDateTime)) +
  mean_grass * mean_forest

az_wf_spcov <- spcov_initial("wave")

az_wf_spglm <- spglm(spglm_formula, data = data_sf, 
                     family = binomial, spcov_initial = az_wf_spcov)
#check spcov_initial

summary(az_wf_spglm)

#model diagnostics

model_cv <- loocv(az_wf_spglm, type = "response")

kable(model_cv, caption = "Leave-One-Out Cross Validation")

kable(AIC(az_wf_spglm), caption = "Model AIC")

```

```{r getpredictions, eval = FALSE}

# Create a grid of points
grid <- st_make_grid(coconino_sf, n = c(20, 20), what = "centers")

# Convert the grid to an sf object
grid_sf <- st_sf(geometry = grid)

# Filter points to keep only those inside Coconino County
coconino_grid <- grid_sf[coconino_sf, ]

# If you need exactly 200 points, you can sample from the resulting grid
coconino_grid_n <- coconino_grid %>% 
  slice_sample(n = 400)

# Plot to verify
plot(st_geometry(coconino_sf))
plot(st_geometry(coconino_grid_n), add = TRUE, col = "red", pch = 20)

prediction_grid <- cbind(st_drop_geometry(coconino_grid_n), st_coordinates(coconino_grid_n)) 

prediction_grid$FireDiscoveryDateTime <- ymd_hm("2023-12-31 12:00")

```

```{r updatepredictgrid, eval=FALSE}

az_prediction_grid <- read_csv(here('data/az_prediction_grid_utm_latlon.csv'))

# az_prediction_grid <- az_prediction_grid %>% select(c(-1))

az_prediction_grid_sf <- st_as_sf(az_prediction_grid, coords = c("X", "Y"), crs = "EPSG:32612")

az_prediction_grid_sf <- st_transform(az_prediction_grid_sf, crs = 4326)

# Extract coordinates
coords_4326 <- st_coordinates(az_prediction_grid_sf)

# Create a dataframe with original coordinates
df <- data.frame(
  lon = coords_4326[,1],
  lat = coords_4326[,2]
)

az_prediction_grid <- cbind(az_prediction_grid, df)

```


```{r predictiongridnatcovs, eval=FALSE}

az_prediction_grid_nat_covs <- read_csv(here('data/az_locations_all.csv'))

az_prediction_grid_nat_covs <- az_prediction_grid_nat_covs %>% select(,-c(1:5))

az_prediction_grid_sf <- cbind(az_prediction_grid_sf, az_prediction_grid_nat_covs)

az_prediction_grid_sf <- st_join(az_prediction_grid_sf, population_data, left = TRUE)



az_prediction_grid_sf$distance_rd_primary <-
  st_distance(az_prediction_grid_sf, az_rd_primary) %>% apply(1, min)

az_prediction_grid_sf$distance_rd_secondary <-
  st_distance(az_prediction_grid_sf, az_rd_secondary) %>% apply(1, min)

az_prediction_grid_sf$distance_rd_4wd <-
  st_distance(az_prediction_grid_sf, az_rd_4wd) %>% apply(1, min)


az_prediction_grid_sf <- az_prediction_grid_sf %>%
  mutate(distance_rd_min_prisec = pmin(distance_rd_primary, 
                                    distance_rd_secondary))

az_prediction_grid_sf <- az_prediction_grid_sf %>%
  mutate(distance_rd_min_all = pmin(distance_rd_primary, 
                                    distance_rd_secondary,
                                    distance_rd_4wd))

az_prediction_grid_sf$distance_rd_min_isprisec <- as.integer(az_prediction_grid_sf$distance_rd_min_all ==                                                     az_prediction_grid_sf$distance_rd_min_prisec)

```




```{r predictionglm}

load(here('data/az_prediction_grid_sf.RData'))

az_wf_predict_spglm <- predict(az_wf_spglm, type = "response", se.fit = T, newdata = az_prediction_grid_sf)

az_prediction_grid_sf$predict_bin_spglm <- az_wf_predict_spglm$fit

predict_plot <- st_transform(az_prediction_grid_sf, crs = 4326)


#generate pixel grid

plot_coconino_sf <- st_transform(coconino_sf, crs = 4326)

# Create a grid of squares
grid <- st_make_grid(plot_coconino_sf, n = c(20, 20), what = "polygons")

# Convert the grid to an sf object
grid_sf <- st_sf(geometry = grid)

# Filter squares to keep only those intersecting with Coconino County
coconino_grid <- grid_sf[plot_coconino_sf, ]

# Spatial join to transfer predicted values from points to grid squares
coconino_grid_with_values <- st_join(coconino_grid, predict_plot)

#plot

ggplot() +
  geom_sf(data = coconino_grid_with_values, aes(fill = predict_bin_spglm), color = NA) +
  geom_sf(data = coconino_sf, fill = NA, color = "black", size = 0.5) +
  scale_fill_viridis_c(option = "B") +
  theme_minimal() +
  labs(title = "Predicted Values in Coconino County",
       fill = "Predicted Value")

```
